A Framework for Substructural Type Systems \thanks{Supported by organization x.}

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] fix up some of the 2-column code
[ writing ] tweaks to Section 4

We capture the form of rules exemplified previously\todo{Previously?} via
\emph{descriptions} of rules.
The key to making these descriptions work is that they only allow syntactic
forms that preserve environments.
These forms are: absence and multiplicity of subterms with the same usage
annotations, absence and multiplicity of subterms with summed usage annotations,
scaling of a subterm, and variable binding.\todo{Switching to Agda}

\paragraph{Premises, Rules, and Systems.}

A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and a type
of conclusion.

A single premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$} constructor.
This allows binding of some variables \AgdaBound{$\Delta$} (with specified
types and usage annotations) and the specification of a conclusion type
\AgdaBound{A}.
The rest of the constructors are descriptions for the first-order bunched
connectives, and are interpreted directly as such.

\ExecuteMetaData[\Syntaxtex]{Premises}
\ExecuteMetaData[\Syntaxtex]{Rule}
\ExecuteMetaData[\Syntaxtex]{System}

@@

As an example, we transcribe the system defined in \cref{fig:lr-comb} into a
description.
We first give the set of types in a standard way (together with base type
\AgdaInductiveConstructor{$\iota$}).

@@

%Then, we give the set of labels.
There is one label for each instantiation of a logical rule, but the labels
contain no further information about subterms or restrictions on the context.

\ExecuteMetaData[\PaperExamplestex]{qlR}

Finally, we specify, for each label, the rest of the information about that
rule {\color{red}(premises and conclusion)}.
We do not have a variable rule, because the variable rule is implicit to any
type system.
{\color{red}(Looks like \cref{fig:lr-comb})}

@@

@@

To specify the meaning of descriptions, we assume some
\ExecuteMetaData[\Interpretationtex]{X}, over which we form one layer of syntax.
The first argument to \AgdaBound{X} is the new variables bound by this layer
of syntax, as exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.
The second argument is the context containing the variables being carried over
from the previous layer.
Notice that this is not, in general, the same as the context from the previous
layer, because the usage annotations may have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and \AgdaInductiveConstructor{\_`$\cdot$\_}.
The third argument is the type of subterm required.

With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
the rest are simply interpretations into bunched combinators.

@@

The interpretation of a \AgdaRecord{Rule} is to check that the rule targets
the desired type and then interpret the rules premises \AgdaBound{ps}.
Notice that the interpretation of the premises is independent of the conclusion
of the rule, which accounts for the difference in type between
\AgdaFunction{$\llbracket$\_$\rrbracket$p} and
\AgdaFunction{$\llbracket$\_$\rrbracket$r}.

@@

The most obvious way to make such an \AgdaBound{X} is to use some existing
\AgdaFunction{OpenFam} on an extended context.
This is what \AgdaFunction{Scope} does --- take the new variables
\AgdaBound{$\Delta$}, concatenate them onto the existing context
\AgdaBound{$\Gamma$}, and pass the extended context onto the judgement
\AgdaBound{T}.

@@

{\color{red}(Forward ref: for now, we could have inlined \texttt{Scope}.)}

It is this method we use to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though that
picture is complicated by variable binding.
A term is either a variable or a use of a logical rule together with terms
for each of the required subterms.

@@

\subsection{Other syntaxes and syntactic forms}

\paragraph{$\mu\tilde\mu$}
We can encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\cite{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kits} is the form
of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\cite{CH00}, we use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it as a natural
deduction system, thus giving a notion of \emph{variable} to the system.

@@

\begin{minipage}[t]{0.5\textwidth}
  \centering
  \ExecuteMetaData[\MuMuTildetex]{Ty}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \centering
  \ExecuteMetaData[\MuMuTildetex]{Conc}
\end{minipage}

>>>>

We take the insights of the previous section and use them to build a
generic framework for posemiring-annotated substructural systems in
Agda. We will first show \emph{descriptions} of systems, which are
comprised of rules that have premises combined using the bunched
combinators. We then show how to construct the Agda data type of
intrinsically well scoped, typed, and resourced terms for any given
system of our framework. We use the prototypical system from
\figref{fig:comb-lr} as a running example. \secref{sec:other-syntaxes}
presents further examples that our framework can express.

We now start to use Agda notation for record and data type
declarations, to emphasise that our framework has been implemented.

\subsection{Descriptions of Systems}

% We capture the form of rules exemplified previously\todo{Previously?} via
% \emph{descriptions} of rules.
% The key to making these descriptions work is that they only allow syntactic
% forms that preserve environments.
% These forms are: absence and multiplicity of subterms with the same usage
% annotations, absence and multiplicity of subterms with summed usage annotations,
% scaling of a subterm, and variable binding.\todo{Switching to Agda}

\paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}

A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and
a type of conclusion. We assume that there is a
$\AgdaBound{Ty} : \AgdaPrimitiveType{Set}$ of types for the system in
scope.

The \AgdaDatatype{Premise} data type describes premises of rules,
using the bunched combinators from the previous section. A single
premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
constructor.  This allows binding of additional variables
\AgdaBound{$\Delta$} (with specified types and usage annotations) and
the specification of a conclusion type \AgdaBound{A} for this premise.
The remaining constructors are descriptions for the first-order
bunched connectives, and will be interpreted directly as such, below.

\ExecuteMetaData[\Syntaxtex]{Premises}

A \AgdaRecord{Rule} is a pair of some \AgdaDatatype{Premises} and a
conclusion. We use an infix arrow as a suggestive notation for rules.

\ExecuteMetaData[\Syntaxtex]{Rule}

Finally, a \AgdaRecord{System} consists of a set of rule labels (i.e.,
constructor names), and for each label a decsription of the
corresponding rule. We use $\rhd$ as infix notation for systems to
associate the label set with the rules.

\ExecuteMetaData[\Syntaxtex]{System}

@@

As an example, we transcribe the system defined in
\figref{fig:lr-comb} into a description.  We first give the set of
types as a data type (together with a base type
\AgdaInductiveConstructor{$\iota$}). We assume that there is a
posemiring \AgdaInductiveConstructor{Ann} in scope for the
annotations.

@@

%Then, we give the set of labels.
There is one label for each instantiation of a logical rule, but the
labels contain no further information about subterms or restrictions
on the context. This will be provided when we associate labels with
\AgdaRecord{Rule}s in a \AgdaRecord{System}.

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{Ty}
  \ExecuteMetaData[\PaperExamplestex]{Side}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{qlR}
\end{minipage}

To build a system, we associate with each label a rule:

@@

Compared to \figref{fig:lr-comb}, modulo the Agda notation, we can see
that the fundamental structure has been preserved: the rules match
one-to-one, and the bunched premises are the same. A major difference
is that we do not include a counterpart to the
\AgdaInductiveConstructor{var} rule in a
\AgdaRecord{System}. Variables are common to all the systems
representable in our framework.

@@

To specify the meaning of descriptions, we assume some \AgdaBound{X} : \AgdaFunction{ExtOpenFam},
% \ExecuteMetaData[\Interpretationtex]{X},
over which we form one layer of syntax, using the function
\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
\AgdaDatatype{Premises} defined below.  The first argument to
\AgdaBound{X} is the new variables bound by this layer of syntax, as
exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
the context containing the variables being carried over from the
previous layer.  Notice that this is not, in general, the same as the
context from the previous layer, because the usage annotations may
have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and
\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
type of subterm required.

With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
the rest are simply interpretations of premises into bunched combinators.

@@

The interpretation of a \AgdaRecord{Rule} checks that the rule targets
the desired type and then interprets the rule's premises \AgdaBound{ps}.
Notice that the interpretation of the premises is independent of the conclusion
of the rule, which accounts for the difference in type between
\AgdaFunction{$\llbracket$\_$\rrbracket$p} and
\AgdaFunction{$\llbracket$\_$\rrbracket$r}.

@@

The most obvious way to make such an \AgdaBound{X} is to use some existing
\AgdaFunction{OpenFam} on an extended context.
We defined \AgdaFunction{Scope} to do this: take the new variables
\AgdaBound{$\Delta$}, concatenate them onto the existing context
\AgdaBound{$\Gamma$}, and pass the extended context onto the judgement
\AgdaBound{T}.

@@

%{\color{red}(Forward ref: for now, we could have inlined \texttt{Scope}.)}

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is where we
use sized types to convince Agda that this type is strictly positive.

@@

\subsection{Other syntaxes and syntactic forms}\label{sec:other-syntaxes}

\paragraph{The system $\mu\tilde\mu$.}
We can encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\cite{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kits} is the form
of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\cite{CH00}, we use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it as a natural
deduction system, thus giving a notion of \emph{variable} to the system.

@@

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\MuMuTildetex]{Ty}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\MuMuTildetex]{Conc}
\end{minipage}
